---
title: "Skills Test 2"
author: "Tom Blackwood"
date: "10/12/2020"
output: 
  # html_document
  rmarkdown:: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load package
library(tidyverse)
library(ggplot2)
library(lubridate)
library(patchwork)
```

## Assignment 2

Fuel data is randomly generated and written to a `.CSV`.  The Seed for the random generation is set ot 72 and the sample size is set to 7,216,516,166.

```{r fuel_gen}
# set seed and sample size
set.seed(72) # change this
sample_size <- 7216516166 # feel free to increase this

# create randomised tibble/data frame
fuel_data <- tibble(
  year = sample(c(2010:2020), min(500, sample_size), replace = TRUE),
  month = sample(c(1:12), min(500, sample_size), replace = TRUE),
  day = sample(c(1:31), min(500, sample_size), replace = TRUE),
  petrol_price = round(rnorm(min(500, sample_size), 110, 1)/100, 2),
  diesel_price = round(rnorm(min(500, sample_size), 120, 1)/100, 2)
)

# write fuel_data to file
write.csv(fuel_data, "fuel_data.csv", row.names=FALSE)
```
The data is then read from the csv and a summary says what's in the data.
```{r fuel_read}
# read file in 
fuel_data2 <- read.csv("fuel_data.csv")

# summarise data in at least one way
head(fuel_data2)
summary(fuel_data2)
```
The data is then articulated to be plotted. 
A new column is created for the `date` so the fuel prices can be pivoted longer to plot both fuels on the same graph.

The `month` number is replaced with the month value and the dates that generated warnings are dropped from the set.

A facet grid plot is created using the `months` for one axis of the grid and the `years` for the other axis. The days of the month are posted along the `x-axis` and the price along the `y-axis`. 

```{r fuel_plot, fig.width=12, fig.height=10}
# plot the data in at least one way
fuel_data2$date <- with(fuel_data2, ymd(paste(year, month, day, sep="-")))
fuel_data3 <- fuel_data2 %>% 
  pivot_longer(cols = petrol_price:diesel_price, names_to = 'fuel', values_to = 'price')

fuel_data3$month <- month(as.Date(fuel_data3$date), label=TRUE)
fuel_data3 <- fuel_data3 %>% drop_na(month)

p1 <- ggplot(fuel_data3, aes(x = day, y = price, colour=fuel)) + 
  geom_line() + 
  facet_grid(year ~ month) + 
  
  ggtitle("Price of Fuel by Month/Year") +
  labs(x = "Day of Month", y = "Price", colour = "Fuel") + 
  
  theme(
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
  )

p1

```