---
title: "Tutorial 6 Functions"
author: "Tom Blackwood"
date: "02/12/2020"
output: 
  # html_document
  rmarkdown:: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
```

## Exercise 1 
### Part 1

Write a function that subtracts one vector from another


```{r subtract_function}

vec_sub <- function(vec_1, vec_2){
  return(vec_1 - vec_2)
}

vec_1 <- c(0,1,2)
vec_2 <- c(6,5,4)

print(vec_sub(vec_1, vec_2))

```

### Part 2

Form a word in a function using vectors

```{r form_word}

make_word <- function(str_vec, is_word = FALSE){
  if (is_word){
    return(str_c(str_vec, collapse = ""))  
  } else {
    return(str_c(str_vec, collapse = " "))  
  }
  
}

some_chars <- c("H", "e", "l", "l", "o")
make_word(some_chars)

some_strings <- c("Hello", "World!", "I", "live", "another", "day!")
make_word(some_strings)

```

### Part 3

Take the code above and add a logic boolean to  tell the function if it's a word or a sentence. 

## Exercise 2

Rewrite the code from exercise 1 to include error checking


```{r form_word_2}

make_word <- function(str_vec, is_word = FALSE){
  str_clean <- c()
  for(char in c(1:length(str_vec))){
    if (is.numeric(str_vec[char])){
      str_clean[char] <- FALSE
      
    } else if (is.character(str_vec[char])){
      str_clean[char] <- TRUE
      
    } else {
      str_clean[char] <- FALSE
    }
  }
  
  if (FALSE %in% str_clean){
    warning("Not a string vector") # In practice, this should be a stop
    
  }else{
    if (is_word){
      if(nchar(str_vec[1]) == 1){
        return(str_c(str_vec, collapse = ""))
        
      } else {
        warning("First character is a word and not a letter.") # In practice, this should be a stop
      }
      
    } else {
      if(nchar(str_vec[1]) > 1){
        return(str_c(str_vec, collapse = " "))
        
      } else {
        warning(str_c(c("First element of the vector is ", str_vec[1], ", could be a word and not a sentence."), collapse=""))
        return(str_c(str_vec, collapse = " "))
      }
    }
  }
}

some_chars <- c("H", "e", "l", "l", "o")
make_word(some_chars, is_word=TRUE)
make_word(some_chars)

some_strings <- c("Hello", "World!", "I", "live", "another", "day!")
make_word(some_strings)
make_word(some_strings, is_word=TRUE)

some_nums <- c(1, 2, 5)
some_stuff <- c(5, "A", TRUE)
some_more <- c(some_chars, some_strings)

make_word(some_nums)
make_word(some_stuff)
make_word(some_more)
make_word(some_more, is_word=TRUE)
```

So, largely it works apart from if you use 2 lists and if you have a list with mixed data.

## Exercise 3 - Sourcefiles

### Part 1 

The cleaning functions from Tutorial 4 are put into a function. The functions are appended with `_local` to distinguish them from functions being written into scripts later in the exercise.

```{r nairn_data}
nairn <- read_tsv("../../data/nairn_raw.tsv")
head(nairn)

clean_data_local <- function(df){
  df[df == "---"] <- "NA"
  df[] <- lapply(df, gsub, pattern = '\\*', replacement = '')
  df[] <- sapply(df[], as.numeric)
  return (df)
}

nairn = clean_data_local(nairn)

nairn %>%
  group_by(`tmax (degC)`) %>%
  count()
```

### Part 2

create a function to read in a file path, clean the data, and returns a cleaned dataset.

```{r read_and_clean_nairn_data}

get_data_local <- function(fp){
  df <- read_tsv(fp)
  df <- clean_data_local(df) # from above
  return (df)
}

nairn2 <- get_data_local("../../data/nairn_raw.tsv") 

nairn2 %>%
  group_by(`tmax (degC)`) %>%
  count()
```

### Part 3

Putting the scripts into an external file marked `my_scripts.R`.

Since my markdown file is in `Tutorials/Tutorial 6`, I have to go up two levels to get to the main folder where `/R/` is to get into `/R/Tutorial 6/my_scripts.R`. 

```{r ext_read_clean_data}

# load packages
source("../../R/Tutorial 6/my_scripts.R")

nairn3 <- get_data("../../data/nairn_raw.tsv") 

nairn3 %>%
  group_by(`tmax (degC)`) %>%
  count()
```

### Part 4

Now the script is going to be used on the `dunstaffnage_raw.tsv` data.

```{r dunstaffnage_raw}

# load packages
source("../../R/Tutorial 6/my_scripts.R")

dunstaffnage <- get_data("../../data/dunstaffnage_raw.tsv") 

dunstaffnage %>%
  group_by(`tmax (degC)`) %>%
  count()

head(dunstaffnage)
summary(dunstaffnage)
```
