---
title: "Tutorial 3 Reading in Data and Date/Strings"
author: "Tom Blackwood"
date: "23/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=10, fig.height=5)
library(tidyverse)
library(ggplot2)
```

## Exercise 1 - Reading using `read_csv()` or `read_tsv()`

My files for this course are kept in a folder named `../intro_to_r/`. The structure for where this markdown file is kept is in `../intro_to_R/tutorials/tutorial */` and my data is held in `../intro_to_R/data/`. The data used is supplied from the data bundle available for the course.

*Very good description of what you're doing. It doesn't matter where you store things really as long as you understand how to access them.*

```{r imports}
csv_data <- read_csv("../../data/student_marks.csv")
head(csv_data)

tsv_data <- read_tsv("../../data/nairn_raw.tsv")
head(tsv_data)

```

## Exercise 2 - Writing using `write_csv()`

writing the storms data to a `.csv`

```{r writing_a_csv}
write_csv(storms, file="../../data/storms.csv")
```

## Exercise 3 - Converting `status` into `factor`

This exercise reads the data from `../data/storms.xlsx`, an excel file in the data folder.

```{r status_to_factor}
library(readxl)
library(forcats)
storms_xl <- read_xlsx("../../data/storms.xlsx")
storms_xl$status <- as_factor(storms_xl$status)
levels(storms_xl$status)

ggplot(storms_xl, aes(status)) +
  geom_bar()

# can also do this within a mutate statement but this is absolutely fine
storms_xl$status_rev <- fct_rev(storms_xl$status) 
ggplot(storms_xl, aes(status_rev)) +
  geom_bar()
```

## Exercise 4 - Student marks `wide` to `long`

This exercise reads the data from `../data/student_marks.csv`, an excel file in the data folder.

```{r student_marks, fig.width=15, fig.height=5}
marks_wide <- read_csv("../../data/student_marks.csv")
marks_long <- pivot_longer(marks_wide, cols = maths:economics, names_to = "subject", values_to = "mark")
head(marks_long)
ggplot(marks_long, aes(x = student, y = mark, fill = subject)) + 
  geom_col(position = "dodge") # Excellent work
```

## Exercise 5 - Markdown vs Scripts

I have only really worked with creating the markdown documents, but I have found that you can't really run script sections in markdown unless you have all the variables and libraries mentioned available in that section, which might have been called earlier in the markdown file.


